---
title: 'CitiBike Rebalancing Study'
author:
  - name: Thida Chong
date: '`r Sys.Date()`'
output: distill::distill_article
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  eval = TRUE,
  echo = TRUE,
  message = FALSE,
  error = FALSE,
  warning = FALSE
)
```
[Columbia University's Center for Spatial Research](https://c4sr.columbia.edu/projects/citibike-rebalancing-study). 

You can download and zoom in on a high resolution pdf of the Spatial Information Design Lab's graphic here: [http://c4sr.spatialinformationdesignlab.org/sites/default/files/Activity_Matrix_Composite.pdf](http://c4sr.spatialinformationdesignlab.org/sites/default/files/Activity_Matrix_Composite.pdf).


```{r}
library(tidyverse)
```

```{r}
savefile <- "data/201909-citibike-tripdata.csv"

if (!file.exists(savefile)) {
  url <- "https://s3.amazonaws.com/tripdata/201909-citibike-tripdata.csv.zip"
  download.file(url = url, destfile = savefile )
  }

df <- read_csv(savefile)
```

renaming variables

```{r}
df <- df %>% rename_with(~ gsub(' ', '_', .) )
```

Remove missing data
Only consider observations with an `end_station_name` that is also used as a `start_station_name`.

```{r}
df <- 
  df %>% 
  filter(
    if_any(contains('station_name'), ~ !is.na(.)),
    end_station_name %in% start_station_name
  )
```

Ccount the number of rides both starting and ending at each station name at each hour of the day, averaged over the number of days in the dataset. 
Pivot data and create new variables:  `start_station_name` and `end_station_name` into long format
The, create variables for day of month (`day`) and hour of day (`hour`)

```{r}
df <- 
  df %>%
  pivot_longer(
    cols = c(start_station_name, end_station_name), 
    names_to = "start_end",
    values_to = "station_name"
  ) %>%
  mutate(
    day  = format( if_else(start_end == "start_station_name", starttime, stoptime), "%d" ),
    hour = format( if_else(start_end == "start_station_name", starttime, stoptime), "%H" )
  ) %>%
  mutate(
    station_name = fct_reorder(station_name, desc(station_name))
  )
```

Create new variables `activity` and `balance`, where `activity` holds the average number of rides or observations at each station name each hour and where `balance` hold the average difference between rides beginning at the station and rides ending at the station.

```{r}
df <- 
  df %>%
  group_by(station_name, hour, .drop = FALSE) %>%
  summarise(
    activity = sum(n()) / n_distinct(day), # average number of rides per day
    balance  = sum(if_else(start_end == "start_station_name", -1, 1)) # difference between rides starting and ending at the station
  ) %>%
  ungroup()
```

scale  data variables to the ranges of your visual variables

```{r}
library(scales)

df <-
  df %>%
  mutate(
    hue = if_else(balance < 0, 50, 200),
    saturation =
      rescale(
        abs(balance),
        from = c(0, 100), #range of 'balance' data
        to   = c(0, 200) #range of saturation 
      ),
    luminance =
      rescale(
        activity,
        from = c(0, max(df$activity)), #range of 'activity' data
        to   = c(0, 100) #range of luminance 
      )
  )
```

map our data onto the visual variables.


```{r}
p <- 
  df %>%
  ggplot() +
  scale_fill_identity() +
  geom_tile(
    mapping = aes( x = hour,
      y = station_name,
      fill = hcl(hue, saturation, luminance)
    ),
    width = 0.95,
    height = 0.95
  ) +
  theme_dark() +
  theme(
    panel.background = element_blank(),
    panel.grid = element_blank(),
    plot.background = element_rect(fill = "#333333"),
    axis.text.x = element_text(color = "#888888", size = 16 / .pt),
    axis.text.y = element_text(color = "#888888", size =  7 / .pt)
  ) +
  labs(x = "", y = "")

# The next line of code will save the graphic as a png
# directory so that you can separately open and zoom in while reviewing it.
ggsave("activity_balance2019.png", plot = p, width = 8, height = 40)

p
```
